Concurrent Programming with Harmony

### 2020-07-19 book/paper.tex
Latest

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language.
\item \autoref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that supports locks and other synchronization primitives.
\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
\autoref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\end{itemize}

@@

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Address $n$ & compute address from $n$ components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert & pop $b$ and check that it is \texttt{True} \\
\hline
AtomicInc & increment the atomic counter of this context \\
\hline
AtomicDec & decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete process variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a process variable onto the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
PushAddress $v$ & push the address of variable $v$ \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop set and push minimum and remainder \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
\hline
Swap & swap the two values on top of the stack \\
\hline
\end{tabular}
\vspace{1em}

>>>>

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language.
\item \autoref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that supports locks and other synchronization primitives.
\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
\autoref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:interrupts} discusses how to handle interrupts to a process,
a problem closely related to synchronization among multiple processes.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\end{itemize}

@@

\chapter{Interrupts}
\label{ch:interrupts}
\index{interrupt}

\begin{figure}
\begin{code}
\harmonysource{trap}
\end{code}
\caption{\harmonylink{code/trap.hny} How to use \textbf{trap}.}
\label{fig:trap}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{trap2}
\end{code}
\caption{\harmonylink{code/trap2.hny} A race condition with interrupts.}
\label{fig:trap}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{trap3}
\end{code}
\caption{\harmonylink{code/trap3.hny} Locks do not work with interrupts.}
\label{fig:trap}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{trap4}
\end{code}
\caption{\harmonylink{code/trap4.hny} Disabling and enabling interrupts.}
\label{fig:trap}
\end{figure}

Processes can be \emph{interrupted}.  An interrupt is a notification of some event
such as a keystroke, a timer expiring, the reception of a network packet, the completion
of a disk operation, and so on.  We distinguish \emph{interrupts} and \emph{exceptions}.
An exception is caused by the process executing an invalid machine instruction such as
divide-by-zero.  An interrupt is caused by some peripheral device.

Harmony allows modeling interrupts using operator \texttt{trap}:
\begin{code}
\textbf{trap} \textit{handler} \textit{argument}
\end{code}
invokes \textit{handler argument} at some later, unspecified time.
Only one of these asynchronous events can be outstanding at a time; a new call to
\textbf{trap} overwrites any outstanding one.
Figure~\ref{fig:trap} gives an example of how \textbf{trap} might be used.
Here, the \textit{main}() process loops until the interrupt has occurred and
the \textit{done} flag has been set.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Address $n$ & compute address from $n$ components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert & pop $b$ and check that it is \texttt{True} \\
\hline
AtomicInc & increment the atomic counter of this context \\
\hline
AtomicDec & decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete process variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a process variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
PushAddress $v$ & push the address of variable $v$ \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop set and push minimum and remainder \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
\hline
\end{tabular}
\vspace{1em}
